Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-06-21 ms.tex
Fix typo
New bacronym?
Add margin note about orbital precession

In this paper we present a rapid and accurate method of sky localization that makes use of Bayesian methods. It differs from existing techniques is several important ways.  In the first place, we fix the intrinsic parameters to their \ac{ML} estimates, as provided by the detection pipeline.  This reduces the dimensionality of the parameter space we need to sample. In addition, the technique takes as its input the \ac{ML} estimates of the extrinsic parameters, instead of the $h(t)$ time series. This makes the likelihood itself much faster to evaluate. Finally, instead of of using \ac{MCMC} or some similar method for statistical sampling, we make use of a deterministic quadrature scheme. We call this technique \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \acp{GW} detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `T' in the acronym, because then it would be called BAYES\nobreakdashes-TART, which would sound ridiculous.}%
. It is unique in that in bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. We expect that \ac{BAYESTAR} will take on a key role in observing \ac{CBC} events in both \ac{GW} and optical channels during the Advanced \ac{LIGO} era.

@@

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \boldsymbol\theta = \begin{array}{l@{}l}
            \left[
            \begin{array}{c}
                \alpha \\
                \delta \\
                D_\mathrm{L} \\
                t_\oplus \\
                u \\
                \psi \\
                \phi_c \\
                \hdashline[1pt/1pt]
                m_1 \\
                m_2 \\
                \mathbf S_1 \\
                \mathbf S_2
            \end{array}
            \right] &
            \begin{array}{l}
                \left.
                \begin{array}{p{4cm}}
                    right ascension \\
                    declination \\
                    luminosity distance \\
                    arrival time at geocenter \\
                    $\cos(\textrm{inclination angle})$ \\
                    polarization angle \\
                    coalescence phase \\
                    \hdashline[1pt/1pt]
                \end{array}
                \quad \right\} \textnormal{extrinsic parameters}, \boldsymbol\theta_\mathrm{ex} \\
                \left.
                \begin{array}{p{4cm}}
                    first component's mass \\
                    second component's mass \\
                    first component's spin \\
                    second component's spin
                \end{array}
                \quad \right\} \textnormal{intrinsic parameters}, \boldsymbol\theta_\mathrm{in}
            \end{array}
        \end{array}
\end{equation}
%
Binary eccentricity is omitted as an intrinisc parameter because tidal interaction of the progenitor stars~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} circularize the orbit long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the effect of tides on the inspiral would only be detectable by an Einstein Telescope-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{NSNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$. In this work we will consider only zero spin.

@@

There are many ways to realize distribution function with these properties, but they tend to be difficult to work with.%
%
\marginpar{Give some examples.}%
%
An alternative is to just discard the one periodic parameter, $\gamma$, entirely. We do not know the exact distribution of the remaining parameters $\rho$ and $\tau$, but we are less liable to make egregious errors by assuming that they are Gaussian than if we assumed that the estimation errors in $\gamma$ were Gaussian. Projecting out $\gamma$ gives us the covariance matrix
%
\begin{equation}
    \cov \left(
        \begin{array}{c}
            \hat{\rho} - \rho \\
            \hat{\tau} - \tau
        \end{array}
    \right) =
    \left(
        \begin{array}{cc}
            1 & 0 \\
            0 & \rho^2/{\omega_\mathrm{rms}}^2
        \end{array}
    \right).
\end{equation}

>>>>

In this paper we present a rapid and accurate method of sky localization that makes use of Bayesian methods. It differs from existing techniques is several important ways.  In the first place, we fix the intrinsic parameters to their \ac{ML} estimates, as provided by the detection pipeline.  This reduces the dimensionality of the parameter space we need to sample. In addition, the technique takes as its input the \ac{ML} estimates of the extrinsic parameters, instead of the $h(t)$ time series. This makes the likelihood itself much faster to evaluate. Finally, instead of of using \ac{MCMC} or some similar method for statistical sampling, we make use of a deterministic quadrature scheme. We call this technique \ac{BAYESTAR}%
%
\marginpar{Probably need to change the backronym. BAYESian TriAngulation and Rapid localization?}
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \acp{GW} detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `T' in the acronym, because then it would be called BAYES\nobreakdashes-TART, which would sound ridiculous.}%
. It is unique in that in bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. We expect that \ac{BAYESTAR} will take on a key role in observing \ac{CBC} events in both \ac{GW} and optical channels during the Advanced \ac{LIGO} era.

@@

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \boldsymbol\theta = \begin{array}{l@{}l}
            \left[
            \begin{array}{c}
                \alpha \\
                \delta \\
                D_\mathrm{L} \\
                t_\oplus \\
                u \\
                \psi \\
                \phi_c \\
                \hdashline[1pt/1pt]
                m_1 \\
                m_2 \\
                \mathbf S_1 \\
                \mathbf S_2
            \end{array}
            \right] &
            \begin{array}{l}
                \left.
                \begin{array}{p{4cm}}
                    right ascension \\
                    declination \\
                    luminosity distance \\
                    arrival time at geocenter \\
                    $\cos(\textrm{inclination angle})$ \\
                    polarization angle \\
                    coalescence phase \\
                    \hdashline[1pt/1pt]
                \end{array}
                \quad \right\} \textnormal{extrinsic parameters}, \boldsymbol\theta_\mathrm{ex} \\
                \left.
                \begin{array}{p{4cm}}
                    first component's mass \\
                    second component's mass \\
                    first component's spin \\
                    second component's spin
                \end{array}
                \quad \right\} \textnormal{intrinsic parameters}, \boldsymbol\theta_\mathrm{in}
            \end{array}
        \end{array}
\end{equation}
%
Binary eccentricity is omitted as an intrinisc parameter because tidal interaction of the progenitor stars~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} circularize the orbit long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the effect of tides on the inspiral would only be detectable by an Einstein Telescope-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{NSNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$. In this work we will consider only zero spin.%
%
\marginpar{We don't need zero spin; we just need negligible orbital precession.}

@@

There are many ways to realize a distribution function with these properties, but they tend to be difficult to work with.%
%
\marginpar{Give some examples.}%
%
An alternative is to just discard the one periodic parameter, $\gamma$, entirely. We do not know the exact distribution of the remaining parameters $\rho$ and $\tau$, but we are less liable to make egregious errors by assuming that they are Gaussian than if we assumed that the estimation errors in $\gamma$ were Gaussian. Projecting out $\gamma$ gives us the covariance matrix
%
\begin{equation}
    \cov \left(
        \begin{array}{c}
            \hat{\rho} - \rho \\
            \hat{\tau} - \tau
        \end{array}
    \right) =
    \left(
        \begin{array}{cc}
            1 & 0 \\
            0 & \rho^2/{\omega_\mathrm{rms}}^2
        \end{array}
    \right).
\end{equation}
